---
title: "Visualising deprivation in Scotland"
description: |
  A look at the 2020 iteration of the Scottish Index of Multiple Deprivation (SIMD)
author:
  - name: David Henderson
    url: https://davidhen.com
    affiliation: Scottish Centre for Administrative Data Research (SCADR)
    affiliation_url: https://www.scadr.ac.uk/
  - name: John MacKintosh
    url: https://johnmackintosh.com/
    affiliation: NHS Highland
    affiliation_url: https://www.nhshighland.scot.nhs.uk/Pages/welcome.aspx
  - name: Nick Bailey
    affiliation: Urban Big Data Centre
    affiliation_url: https://www.ubdc.ac.uk 
date: "`r Sys.Date()`"
output: distill::distill_article
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  R.options = list(width = 70)
  )
```

```{r, packages, message=FALSE, results='hide'}
library(tidyverse)
library(ggthemes)
library(ggExtra)
library(scales)
library(readxl)
library(httr)
library(janitor)
library(here)
here()

theme_set(theme_minimal(base_size = 26, base_family = "Roboto") +
            theme(panel.grid.minor = element_blank()))
```


```{r, data_and_clean, warning=FALSE, message=FALSE, results='hide'}
#The third sheet in the csv is the one with the data that I want. 
url <- "https://www.gov.scot/binaries/content/documents/govscot/publications/statistics/2020/01/scottish-index-of-multiple-deprivation-2020-data-zone-look-up-file/documents/scottish-index-of-multiple-deprivation-data-zone-look-up/scottish-index-of-multiple-deprivation-data-zone-look-up/govscot%3Adocument/SIMD_2020_Datazone_lookup_tool.xlsx"

GET(url, write_disk(tf <- tempfile(fileext = ".xlsx")))

simd <- read_excel(tf, 3L)

simd %>% 
  #convert all column names to lower and snake case
  #I should come up with better names here but am being lazy
  janitor::clean_names() %>% 
  #keep only the columns of interest
  select(l_aname, population, simd2020_decile) %>% 
  #clean_names didn't quite have the desired effct for la name - sort it here
  rename(la_name = l_aname) %>% 
  #Now group by council and simd decile
  group_by(la_name) %>% 
  #Add a variable which counts the number of datazones
  #(in each council)
  mutate(n_dz = n()) %>% 
  #now group by la and SIMD decile
  group_by(la_name, simd2020_decile) %>%
  #Now calculate pct of datazones in each decile per LA
  #P.S. don't multiply by 100 here as ggplot will do that for us later
  mutate(pct = n()/n_dz) %>% 
  #easier to read (for a human it breaking the pipe here for a look)
  arrange(la_name, simd2020_decile) %>% 
  ungroup %>% 
  #keep the first row of data for each council and decile 
  #(basically a dedup)
  distinct(la_name, simd2020_decile, .keep_all = TRUE) %>% 
  #now factorise the SIMD variable for better plotting options
  mutate(simd2020_decile = factor(simd2020_decile,
                                  levels = c(1:10),
                                  labels = c("1", "2", "3", "4", "5",
                                             "6", "7", "8", "9", "10"))) %>% 
  #drop the now unnecessary population variable
  select(-population) -> clean_simd
```


# Introduction

Yada Yada Yad, new version came out so we wanted to play

## What is SIMD?

Brief description

## Why do we use it?

Helps stratify use of services etc. etc. 

## Any good examples?

Link (and maybe copy a plot) to a paper using SIMD which has had high impact

# SIMD 2020

Data comed from the Scottish Government's SIMD 2020 webpage [here](insert link address) 

## Deprivation by Local Authority

```{r la_simd, fig.width=20, fig.height=18}
clean_simd %>% 
  ggplot(aes(simd2020_decile, pct, fill = simd2020_decile)) +
  geom_col() +
  scale_fill_ptol(guide = guide_legend(nrow = 1),
                  labels = c("1-most deprived", "2", "3", "4", "5",
                             "6", "7", "8", "9", "10-most affluent")) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  facet_wrap(~la_name, strip.position = "bottom") +
  theme(legend.position = "bottom",
        axis.text.x = element_blank()) +
  labs(title = "Percentage of datazones in SIMD 2020 deciles",
       subtitle = "by Local Authority",
       x = "",
       y = "",
       fill = "SIMD decile", 
       caption = "Data from https://www.gov.scot/publications/scottish-index-of-multiple-deprivation-2020-data-zone-look-up/") -> simd_la
simd_la
```

Here we see the variation across the country blah blah blah


Amazing difference between g and E


```{r fig.width=16, fig.height=9}
clean_simd %>% 
  filter(la_name %in% c("Glasgow City", "City of Edinburgh")) %>% 
  ggplot(aes(simd2020_decile, pct, fill = simd2020_decile)) +
  geom_col() +
  scale_fill_ptol(guide = guide_legend(nrow = 1),
                  labels = c("1-most deprived", "2", "3", "4", "5",
                             "6", "7", "8", "9", "10-most affluent")) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  facet_wrap(~la_name, strip.position = "bottom") +
  theme(legend.position = "bottom",
        axis.text.x = element_blank(),
        plot.caption = element_text(size = 12)) +
  labs(title = "Percentage of datazones in SIMD 2020 deciles",
       subtitle = "A tale of two cities",
       x = "",
       y = "",
       fill = "SIMD decile", 
       caption = "Data from https://www.gov.scot/publications/scottish-index-of-multiple-deprivation-2020-data-zone-look-up/") -> simd_la_2cities
simd_la_2cities
```


## SIMD domains by Local Authority

One of John's plots in here, sorry John but I think the black will need to go to fit in with the overall aesthetic, a pure coincidence honest.....

```{r, fig.width=22, fig.height=20}
simd %>% 
  select(DZ, DZname,IZname, LAname,SIMD2020_Access_Domain_Rank,
         SIMD2020_Crime_Domain_Rank,SIMD2020_Decile,
         SIMD2020_Income_Domain_Rank, SIMD2020_Employment_Domain_Rank,
         SIMD2020_Education_Domain_Rank, SIMD2020_Health_Domain_Rank,
         SIMD2020_Housing_Domain_Rank, SIMD2020_Income_Domain_Rank) %>% 
  rename(Access = SIMD2020_Access_Domain_Rank,
         Crime = SIMD2020_Crime_Domain_Rank,
         Decile = SIMD2020_Decile,
         Income = SIMD2020_Income_Domain_Rank, 
         Employment = SIMD2020_Employment_Domain_Rank,
         Education = SIMD2020_Education_Domain_Rank,
         Health = SIMD2020_Health_Domain_Rank,
         Housing = SIMD2020_Housing_Domain_Rank, 
         Income = SIMD2020_Income_Domain_Rank) %>% 
  tidyr::gather(key = "metric", value = "rank",-c(DZ, DZname, IZname, LAname, Decile)) -> ranked
ranked <- ranked %>% 
  mutate(rescaled_rank = rescale(rank))

ranked %>% 
  ggplot(aes(metric,y = rank)) +
  geom_point(aes(colour = rank), position = position_jitter(width = 0.2),alpha = 0.2) +
  scale_colour_viridis_c(option = "plasma", guide = guide_colourbar(barwidth = 30)) +
  facet_wrap(~ LAname, ncol = 6) +
  coord_flip() +
  theme(legend.position = "bottom",
        axis.text.x = element_blank()) +
  labs(title = "SIMD 2020 Domains by Data Zone",
        subtitle = "Grouped by Local Authority",
       y = "Domain",
       x = "", 
       colour = "Rank",
     caption = "@_johnmackintosh \n Data from https://www.gov.scot/publications/scottish-index-of-multiple-deprivation-2020-data-zone-look-up/")
```



## Deprivation by Health Board

## Other plots

Maybe something about local and national share here. I don't like the SG version so could maybe count the ranks up then list local authorities by highest to lowest rank?

# Other resources


# Acknowledgments {.appendix}

This is a place to recognize people and institutions. It may also be a good place
to acknowledge and cite software that makes your work possible.




